---
title: "Description and usage of Spectra objects"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Description and usage of Spectra object}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{Spectra}
    %\VignetteDepends{Spectra,mzR,BiocStyle,msdata,magrittr}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("Spectra")`<br />
**Authors**: `r packageDescription("Spectra")[["Author"]] `<br />
**Last modified:** `r file.info("Spectra.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Spectra)
library(BiocStyle)
```

# Introduction

The `Spectra` package provides a scalable and flexible infrastructure to
represent, retrieve and handle mass spectrometry (MS) data. The `Spectra` object
provides the user with a single standardized interface to access and manipulate
MS data while supporting, through the concept of exchangeable *backends*, a
large variety of different ways to store and retrieve mass spectrometry
data. Such backends range from mzML/mzXML/CDF files, simple flat files, or
database systems.

This vignette provides general examples and descriptions for the `Spectra`
package. Additional information and tutorials can be found
[here](https://jorainer.github.io/SpectraTutorials/).


# Installation

The package can be installed with the `BiocManager` package. To
install `BiocManager` use `install.packages("BiocManager")` and, after that,
`BiocManager::install("Spectra")` to install `Spectra`.


# General usage

Mass spectrometry data in `Spectra` objects can be thought of as a list of
individual spectra, with each spectrum having a set of variables associated with
it. Besides *core* spectra variables (such as MS level or retention time)
an arbitrary number of optional variables can be assigned to a spectrum. The
core spectra variables all have their own accessor method and it is
guaranteed that a value is returned by it (or `NA` if the information is not
available). The core variables and their data type are (alphabetically
ordered):

- *acquisitionNum* `integer(1)`: the index of acquisition of a spectrum during a
  MS run.
- *centroided* `logical(1)`: whether the spectrum is in profile or centroid
  mode.
- *collisionEnergy* `numeric(1)`: collision energy used to create an MSn
  spectrum.
- *dataOrigin* `character(1)`: the *origin* of the spectrum's data, e.g. the
  mzML file from which it was read.
- *dataStorage* `character(1)`: the (current) storage location of the spectrum
  data. This value depends on the backend used to handle and provide the
  data. For an *in-memory* backend like the `MsBackendDataFrame` this will be
  `"<memory>"`, for an on-disk backend such as the `MsBackendHdf5Peaks` it will
  be the name of the HDF5 file where the spectrum's peak data is stored.
- *intensity* `numeric`: intensity values for the spectrum's peaks.
- *isolationWindowLowerMz* `numeric(1)`: lower m/z for the isolation window in
  which the (MSn) spectrum was measured.
- *isolationWindowTargetMz* `numeric(1)`: the target m/z for the isolation
  window in which the (MSn) spectrum was measured.
- *isolationWindowUpperMz* `numeric(1)`: upper m/z for the isolation window in
  which the (MSn) spectrum was measured.
- *msLevel* `integer(1)`: the MS level of the spectrum.
- *mz* `numeric`: the m/z values for the spectrum's peaks.
- *polarity* `integer(1)`: the polarity of the spectrum (`0` and `1`
  representing negative and positive polarity, respectively).
- *precScanNum* `integer(1)`: the scan (acquisition) number of the precursor for
  an MSn spectrum.
- *precursorCharge* `integer(1)`: the charge of the precursor of an MSn
  spectrum.
- *precursorIntensity* `numeric(1)`: the intensity of the precursor of an MSn
  spectrum.
- *precursorMz* `numeric(1)`: the m/z of the precursor of an MSn spectrum.
- *rtime* `numeric(1)`: the retention time of a spectrum.
- *scanIndex* `integer(1)`: the index of a spectrum within a (raw) file.
- *smoothed* `logical(1)`: whether the spectrum was smoothed.

For details on the individual variables and their getter/setter function see the
help for `Spectra` (`?Spectra`). Also note that these variables are suggested,
but not required to characterize a spectrum. Also, some only make sense for MSn,
but not for MS1 spectra.


## Creating `Spectra` objects

The simplest way to create a `Spectra` object is by defining a `DataFrame` with
the corresponding spectra data (using the corresponding spectra variable names
as column names) and passing that to the `Spectra` constructor function. Below
we create such an object for a set of 3 spectra providing their MS level,
polarity but also additional annotations such as their ID in
[HMDB](http://hmdb.ca) (human metabolome database) and their name. The m/z and
intensity values for each spectrum have to be provided as a `list` of `numeric`
values.

```{r spectra-dataframe, message = FALSE}
library(Spectra)

spd <- DataFrame(
    msLevel = c(2L, 2L, 2L),
    polarity = c(1L, 1L, 1L),
    id = c("HMDB0000001", "HMDB0000001", "HMDB0001847"),
    name = c("1-Methylhistidine", "1-Methylhistidine", "Caffeine"))

## Assign m/z and intensity values.
spd$mz <- list(
    c(109.2, 124.2, 124.5, 170.16, 170.52),
    c(83.1, 96.12, 97.14, 109.14, 124.08, 125.1, 170.16),
    c(56.0494, 69.0447, 83.0603, 109.0395, 110.0712,
      111.0551, 123.0429, 138.0662, 195.0876))
spd$intensity <- list(
    c(3.407, 47.494, 3.094, 100.0, 13.240),
    c(6.685, 4.381, 3.022, 16.708, 100.0, 4.565, 40.643),
    c(0.459, 2.585, 2.446, 0.508, 8.968, 0.524, 0.974, 100.0, 40.994))

sps <- Spectra(spd)
sps
```

Alternatively, it is possible to import spectra data from mass spectrometry raw
files in mzML/mzXML or CDF format. Below we create a `Spectra` object from two
mzML files and define to use a `MsBackendMzR` backend to *store* the data (note
that this requires the `r Biocpkg("mzR")` package to be installed). This
backend, specifically designed for raw MS data, keeps only a subset of spectra
variables in memory while reading the m/z and intensity values from the original
data files only on demand. See section [Backends](#backends) for more details
on backends and their properties.

```{r spectra-msbackendmzr, message = FALSE}
fls <- dir(system.file("sciex", package = "msdata"), full.names = TRUE)
sps_sciex <- Spectra(fls, backend = MsBackendMzR())
sps_sciex
```

The `Spectra` object `sps_sciex` allows now to access spectra data from 1862 MS1
spectra and uses `MsBackendMzR` as backend (the `Spectra` object `sps` created
in the previous code block uses the default `MsBackendDataFrame`).


## Accessing spectrum data

As detailed above `Spectra` objects can contain an arbitrary number of
properties of a spectrum (so called *spectra variables*). The available
variables can be listed with the `spectraVariables` method:

```{r spectravariables}
spectraVariables(sps)
spectraVariables(sps_sciex)
```

The two `Spectra` contain a different set of variables: besides `"msLevel"`,
`"polarity"`, `"id"` and `"name"`, that were specified for the `Spectra` object
`sps`, it contains more variables such as `"rtime"`, `"acquisitionNum"` and
`"scanIndex"`. These are part of the *core variables* defining a spectrum and
for all of these accessor methods exist. Below we use `msLevel` and `rtime` to
access the MS levels and retention times for the spectra in `sps`.

```{r mslevel-sps}
msLevel(sps)
rtime(sps)
```

We did not specify retention times for the spectra in `sps` thus `NA` is
returned for them. The `Spectra` object `sps_sciex` contains many more
variables, all of which were extracted from the mzML files. Below we extract the
retention times for the first spectra in the object.

```{r rtime-spssciex}
head(rtime(sps_sciex))
```

Note that in addition to the accessor functions it is also possible to use `$`
to extract a specific spectra variable. To extract the name of the compounds in
`sps` we can use `sps$name`, or, to extract the MS levels `sps$msLevel`.

```{r dollar-extract}
sps$name
sps$msLevel
```

We could also replace specific spectra variables using either the dedicated
method or `$`. Below we specify that all spectra in `sps` represent centroided
data.

```{r dollar-set}
sps$centroided <- TRUE

centroided(sps)
```

The `$` operator can also be used to add arbitrary new spectra variables to a
`Spectra` object. Below we add the SPLASH key to each of the spectra.

```{r new-spectra-variable}
sps$splash <- c(
    "splash10-00di-0900000000-037d24a7d65676b7e356",
    "splash10-00di-0900000000-03e99316bd6c098f5d11",
    "splash10-000i-0900000000-9af60e39c843cb715435")
```

This new spectra variable will now be listed as an additional variable in the
result of the `spectraVariables` function and we can directly access its
content with `sps$splash`.

Each spectrum can have a different number of mass peaks, each consisting of a
mass-to-charge (m/z) and associated intensity value. These can be extracted with
the `mz` or `intensity` functions, each of which return a `list` of `numeric`
values.

```{r mz-intensity}
mz(sps)
intensity(sps)
```

Peak data can also be extracted with the `peaksData` function that returns a
list of matrices, each with columns `"mz"` and `"intensity"` containing the m/z
and intensity values for one spectrum.

```{r peaks}
pks <- peaksData(sps)
pks[[1]]
```

Note that we would get the same result by using the `as` method to coerce a
`Spectra` object to a `list` or `SimpleList`:

```{r as}
as(sps, "SimpleList")
```

The `spectraData` function returns a `DataFrame` with the full data for each
spectrum (except m/z and intensity values), or with selected spectra variables
(which can be specified with the `columns` parameter). Below we extract the
spectra data for variables `"msLevel"`, `"id"` and `"name"`.

```{r spectradata}
spectraData(sps, columns = c("msLevel", "id", "name"))
```

`Spectra` are one-dimensional objects storing spectra, even from different files
or samples, in a single list. Specific variables have thus to be used to define
the originating file from which they were extracted or the sample in which they
were measured. The *data origin* of each spectrum can be extracted with the
`dataOrigin` function. For `sps`, the `Spectra` created from a `DataFrame`, this
will be `NA` because we did not specify the data origin:

```{r dataOrigin-sps}
dataOrigin(sps)
```

`dataOrigin` for `sps_sciex`, the `Spectra` which was initialized with data
from mzML files, in contrast, returns the originating file names:

```{r dataOrigin-sciex}
head(basename(dataOrigin(sps_sciex)))
```

The current data storage location of a spectrum can be retrieved with the
`dataStorage` variable, which will return an arbitrary string for `Spectra` that
use an in-memory backend or the file where the data is stored for on-disk
backends:

```{r dataStorage}
dataStorage(sps)
head(basename(dataStorage(sps_sciex)))
```


## Filtering, subsetting and merging

Apart from *classical* subsetting operations such as `[` and `split`, a set of
filter functions are defined for `Spectra` objects (for detailed help please see
the `?Spectra` help):

- `filterAcquisitionNum`: retain spectra with certain acquisition numbers.
- `filterDataOrigin`: subset to spectra from specific origins.
- `filterDataStorage`: subset to spectra from certain data storage files.
- `filterEmptySpectra`: remove spectra without mass peaks.
- `filterMzRange`: subset spectra keeping only peaks with an m/z within the
  provided m/z range.
- `filterMzValues`: subset spectra keeping only peaks matching provided m/z
  value(s).
- `filterIsolationWindow`: keep spectra with the provided `mz` in their
  isolation window (m/z range).
- `filterMsLevel`: filter by MS level.
- `filterPolarity`: filter by polarity.
- `filterPrecursorMz`: retain (MSn) spectra with a precursor m/z within the
  provided m/z range.
- `filterPrecursorCharge`: retain (MSn) spectra with speified
  precursor charge(s).
- `filterPrecursorScan`: retain (parent and children) scans of an acquisition
  number.
- `filterRt`: filter based on retention time ranges.

In the example below we select all spectra measured in the second mzML file and
subsequently filter them to retain spectra measured between 175 and 189 seconds
in the measurement run.

```{r filterfile-filterrt}
fls <- unique(dataOrigin(sps_sciex))
file_2 <- filterDataOrigin(sps_sciex, dataOrigin = fls[2])
length(file_2)

sps_sub <- filterRt(file_2, rt = c(175, 189))
length(sps_sub)
```

In addition, `Spectra` support also subsetting with `[`. Below we perform the
filtering above with `[` -based subsetting.

```{r subset-square-bracket}
sps_sciex[sps_sciex$dataOrigin == fls[2] &
          sps_sciex$rtime >= 175 &
          sps_sciex$rtime <= 189]
```

The equivalent using filter function is shown below, with the added
benefit that the filtering is recorded in the processing slot.

```{r subset-filter-pipes}
library("magrittr")
sps_sciex %>%
    filterDataOrigin(fls[2]) %>%
    filterRt(c(175, 189))
```

Note that the use of the filter functions might be more efficient for
some backends, depending on their implementation, (e.g. database-based
backends could *translate* the filter function into a SQL condition to
perform the subsetting already within the database).

Multiple `Spectra` objects can also be combined into a single `Spectra` with the
`c` or the `concatenateSpectra` function. The resulting `Spectra` object will
contain an union of the spectra variables of the individual objects. Below we
combine the `Spectra` object `sps` with an additional object containing another
MS2 spectrum for Caffeine.

```{r caf}
caf_df <- DataFrame(msLevel = 2L, name = "Caffeine",
                    id = "HMDB0001847",
                    instrument = "Agilent 1200 RRLC; Agilent 6520 QTOF",
                    splash = "splash10-0002-0900000000-413259091ba7edc46b87",
                    centroided = TRUE)
caf_df$mz <- list(c(110.0710, 138.0655, 138.1057, 138.1742, 195.9864))
caf_df$intensity <- list(c(3.837, 32.341, 0.84, 0.534, 100))

caf <- Spectra(caf_df)
```

Next we combine the two objects.

```{r combine}
sps <- concatenateSpectra(sps, caf)
sps
```

The resulting object contains now the data for all 4 MS2 spectra and an union of
all spectra variables from both objects.

```{r merge-spectravariables}
spectraVariables(sps)
```

The second object had an additional spectra variable *instrument* that was not
present in `sps` and all the spectra in this object will thus get a value of
`NA` for this variable.

```{r merge-add-column}
sps$instrument
```

Sometimes not all spectra variables might be required (e.g. also because many of
them are empty). This might be specifically interesting also for `Spectra`
containing the data from very large experiments, because it can significantly
reduce the object's size in memory. In such cases the `selectSpectraVariables`
function can be used to retain only specified spectra variables.


## Data manipulations

Some analyses require manipulation of the mass peak data (i.e. the m/z and/or
intensity values). One example would be to remove all peaks from a spectrum that
have an intensity lower than a certain threshold. Below we perform such an
operation with the `replaceIntensitiesBelow` function to replace peak
intensities below 10 in each spectrum in `sps` with a value of 0.

```{r replaceintensities}
sps_rep <- replaceIntensitiesBelow(sps, threshold = 10, value = 0)
```

As a result intensities below 10 were set to 0 for all peaks.

```{r replaceintensities-intensity}
intensity(sps_rep)
```

Zero-intensity peaks (and peaks with missing intensities) can then be removed
with the `filterIntensity` function specifying a lower required intensity level
or optionally also an upper intensity limit.

```{r clean}
sps_rep <- filterIntensity(sps_rep, intensity = c(0.1, Inf))
```

```{r clean-intensity}
intensity(sps_rep)
```

The `filterIntensity` supports also a user-provided function to be passed with
parameter `intensity` which would allow e.g. to remove peaks smaller than the
median peak intensity of a spectrum. See examples in the `?filterIntensity` help
page for details.

Note that any data manipulations on `Spectra` objects are not immediately
applied to the peak data. They are added to a so called *processing queue* which
is applied each time peak data is accessed (with the `peaksData`, `mz` or
`intensity` functions). Thanks to this processing queue data manipulation
operations are also possible for *read-only* backends (e.g. mzML-file based
backends or database-based backends). The information about the number of such
processing steps can be seen below (next to *Lazy evaluation queue*).

```{r processing-queue}
sps_rep
```

It is possible to add also custom functions to the processing queue of a
`Spectra` object. Such a function must take a peaks matrix as its first
argument, have `...` in the function definition and must return a peaks matrix
(a peaks matrix is a numeric two-column matrix with the first column containing
the peaks' m/z values and the second the corresponding intensities). Below we
define a function that divides the intensities of each peak by a value which can
be passed with argument `y`.

```{r define-function}
## Define a function that takes a matrix as input, divides the second
## column by parameter y and returns it. Note that ... is required in
## the function's definition.
divide_intensities <- function(x, y, ...) {
    x[, 2] <- x[, 2] / y
    x
}

## Add the function to the procesing queue
sps_2 <- addProcessing(sps_rep, divide_intensities, y = 2)
sps_2
```

Object `sps_2` has now 3 processing steps in its lazy evaluation queue. Calling
`intensity` on this object will now return intensities that are half of the
intensities of the original objects `sps`.

```{r custom-processing}
intensity(sps_2)
intensity(sps_rep)
```

Alternatively we could define a function that returns the maximum peak from each
spectrum:

```{r return-max-peak}
max_peak <- function(x, ...) {
    x[which.max(x[, 2]), , drop = FALSE]
}

sps_2 <- addProcessing(sps_rep, max_peak)
lengths(sps_2)
intensity(sps_2)
```

Each spectrum in `sps_2` thus contains only a single peak. The parameter
`spectraVariables` of the `addProcessing` function allows in addition to define
spectra variables that should be passed (in addition to the peaks matrix) to the
user-provided function. This would enable for example to calculate *neutral
loss* spectra from a `Spectra` by subtracting the precursor m/z from each m/z of
a spectrum. Our tool example does not have precursor m/z values defined, thus we
first set them to arbitrary values. Then we define a function `neutral_loss`
that subtracts the precursor m/z from each peak m/z removing also peaks with an
m/z that is larger or equal to the precursor m/z. Note that, in order to be
able to access the precursor m/z of the spectrum within our function, we have to
add a parameter to the function that has the same name as the spectrum variable
we want to access (in our case `precursorMz`).

```{r set-precursor-mz}
sps_rep$precursorMz <- c(10, 20, 30, 40)

neutral_loss <- function(x, precursorMz, ...) {
    x <- x[x[, "mz"] < precursorMz, , drop = FALSE]
    x[, "mz"] <- abs(x[, "mz"] - precursorMz)
    x
}
```

We have then to call `addProcessing` with `spectraVariables = "precursorMz"` to
specify that this spectra variable is passed along to our function.

```{r neutral-loss}
sps_3 <- addProcessing(sps_rep, neutral_loss, spectraVariables = "precursorMz")
mz(sps_rep)
mz(sps_3)
```

As we can see, the precursor m/z was subtracted from each m/z of the respective
spectrum. A better version of the function, that only calculates neutral loss
spectra for MS level 2 spectra would be the `neutral_loss` function below. Since
we are accessing also the spectrum's MS level we have to call `addProcessing`
adding also the spectra variable `msLevel` to the `spectraVariables`
parameter. Note however that the `msLevel` spectra variable **is by default
renamed** to `spectrumMsLevel` prior passing it to the function. We have thus to
use a parameter called `spectrumMsLevel` in the `neutral_loss` function instead
of `msLevel`.

```{r neutral-loss2}
neutral_loss <- function(x, spectrumMsLevel, precursorMz, ...) {
    if (spectrumMsLevel == 2L)
        x[, "mz"] <- x[, "mz"] - precursorMz
    x
}
sps_3 <- addProcessing(sps_rep, neutral_loss,
                       spectraVariables = c("msLevel", "precursorMz"))
mz(sps_3)
```

Using the same concept it would also be possible to provide any
spectrum-specific user-defined value to the processing function. This variable
could simply be added first as a new spectra variable to the `Spectra` object
and then this variable could be passed along to the function in the same way we
passed the precursor m/z to our function above.

Since all data manipulations above did not change the original intensity or m/z
values, it is possible to *restore* the original data. This can be done with the
`reset` function which will empty the lazy evaluation queue and call the `reset`
method on the storage backend. Below we call `reset` on the `sps_2` object and
hence restore the data to its original state.

```{r reset}
sps_2_rest <- reset(sps_2)

intensity(sps_2_rest)
intensity(sps)
```

Finally, for `Spectra` that use a *writeable* backend, such as the
`MsBackendDataFrame` or `MsBackendHdf5Peaks`, it is possible to apply the
processing queue to the peak data and write that back to the data storage with
the `applyProcessing` function. Below we use this to make all data manipulations
on peak data of the `sps_rep` object persistent.

```{r applyProcessing}
length(sps_rep@processingQueue)

sps_rep <- applyProcessing(sps_rep)
length(sps_rep@processingQueue)
sps_rep
```

Before `applyProcessing` the lazy evaluation queue contained 2 processing steps,
which were then applied to the peak data and *written* to the data storage. Note
that calling `reset` **after** `applyProcessing` can no longer *restore* the
data.


## Comparing spectra and other functions

Spectra can be compared with the `compareSpectra` function, that allows to
calculate similarities between spectra using a variety of methods. However,
peaks from the compared spectra have to be first matched before similarities can
be calculated. `compareSpectra` uses by default the [joinPeaks()] function from
the `r Biocpkg("MsCoreUtils")` package but supports also other mapping functions
to be passed with the `MAPFUN` parameter (see `?joinPeaks` man page in
`MsCoreUtils` for more details). The similarity calculation function can be
specified with the `FUN` parameter and defaults to [ndotproduct()], the
normalized dot-product. Below we calculate pairwise similarities between all
spectra in `sps` accepting a 50 ppm difference of peaks' m/z values for being
considered matching.

```{r comparespectra}
compareSpectra(sps, ppm = 50)
```

The resulting matrix represents the result from the pairwise comparison. As
expected, the first two and the last two spectra are similar, albeit only
moderately while the spectra from 1-Methylhistidine don't share any similarity
with those of Caffeine.


## Plotting `Spectra`

The `Spectra` package provides the following functions to visualize spectra
data:
- `plotSpectra`: plot each spectrum in `Spectra` in its own panel.
- `plotSpectraOverlay`: plot multiple spectra into the **same** plot.

Below we use `plotSpectra` to plot the 4 spectra from the `sps` object using
their names (as provided in spectra variable `"name"`) as plot titles.

```{r plotspectra, fig.width = 8, fig.height = 8}
plotSpectra(sps, main = sps$name)
```

It is also possible to label individual peaks in each plot. Below we use the m/z
value of each peak as its label. In the example we define a function that
accesses information from each spectrum (`z`) and returns a `character` for each
peak with the text that should be used as label. Parameters `labelSrt`,
`labelPos` and `labelOffset` define the rotation of the label text and its
position relative to the x and y coordinates of the peak.

```{r plotspectra-label, fig.width = 8, fig.height = 8}
plotSpectra(sps, main = sps$name,
            labels = function(z) format(mz(z)[[1L]], digits = 4),
            labelSrt = -30, labelPos = 2, labelOffset = 0.1)
```

These plots are rather busy and for some peaks the m/z values are
overplotted. Below we define a *label function* that will only indicate the m/z
of peaks with an intensity higher than 30.

```{r plotspectra-label-int, fig.width = 8, fig.height = 8}
mzLabel <- function(z) {
    z <- peaksData(z)[[1L]]
    lbls <- format(z[, "mz"], digits = 4)
    lbls[z[, "intensity"] < 30] <- ""
    lbls
}
plotSpectra(sps, main = sps$name, labels = mzLabel,
            labelSrt = -30, labelPos = 2, labelOffset = 0.1)
```

Sometimes it might be of interest to plot multiple spectra into the **same**
plot (e.g. to directly compare peaks from multiple spectra). This can be done
with `plotSpectraOverlay` which we use below to create an *overlay-plot* of our
4 example spectra, using a different color for each spectrum.

```{r plotspectraoverlay, fig.width = 6, fig.height = 6}
cols <- c("#E41A1C80", "#377EB880", "#4DAF4A80", "#984EA380")
plotSpectraOverlay(sps, lwd = 2, col = cols)
legend("topleft", col = cols, legend = sps$name, pch = 15)
```

Lastly, `plotSpectraMirror` allows to plot two spectra against each other as a
*mirror plot* which is ideal to visualize spectra comparison results. Below we
plot a spectrum of 1-Methylhistidine against one of Caffeine.

```{r plotspectramirror, fig.width = 6, fig.height = 6}
plotSpectraMirror(sps[1], sps[3])
```

The upper panel shows the spectrum from 1-Methylhistidine, the lower the one of
Caffeine. None of the peaks of the two spectra match. Below we plot the two
spectra of 1-Methylhistidine and the two of Caffeine against each other matching
peaks with a `ppm` of 50.

```{r plotspectramirror-ppm, fig.width = 12, fig.height = 6}
par(mfrow = c(1, 2))
plotSpectraMirror(sps[1], sps[2], main = "1-Methylhistidine", ppm = 50)
plotSpectraMirror(sps[3], sps[4], main = "Caffeine", ppm = 50)
```

See also `?plotSpectra` for more plotting options and examples.


## Exporting spectra

Spectra data can be exported with the `export` method. This method takes the
`Spectra` that is supposed to be exported and the backend (parameter `backend`)
which should be used to export the data and additional parameters for the export
function of this backend. The backend thus defines the format of the exported
file. Note however that not all `MsBackend` classes might support data export.
The backend classes currently supporting data export and its format are:
- `MsBackendMzR` (`Spectra` package): export data in *mzML* and *mzXML*
  format. Can not export all custom, user specified spectra variables.
- `MsBackendMgf`
  ([`MsBackendMgf`](https://github.com/RforMassSpectrometry/MsBackendMgf)
  package): exports data in *Mascot Generic Format* (mgf). Exports all spectra
  variables as individual spectrum fields in the mgf file.

In the example below we use the `MsBackendMzR` to export all spectra from the
variable `sps` to an mzML file. We thus pass the data, the backend that should
be used for the export and the file name of the result file (a temporary file)
to the `export` function (see also the help page of the `export,MsBackendMzR`
function for additional supported parameters).

```{r export}
fl <- tempfile()
export(sps, MsBackendMzR(), file = fl)
```

To evaluate which of the spectra variables were exported, we load the exported
data again and identify spectra variables in the original file which could not
be exported (because they are not defined variables in the mzML standard).

```{r export-import}
sps_im <- Spectra(backendInitialize(MsBackendMzR(), fl))
spectraVariables(sps)[!spectraVariables(sps) %in% spectraVariables(sps_im)]
```

These additional variables were thus not exported. How data export is performed
and handled depends also on the used backend. The `MsBackendMzR` for example
exports all spectra by default to a single file (specified with the `file`
parameter), but it allows also to specify for each individual spectrum in the
`Spectra` to which file it should be exported (parameter `file` has thus to be
of length equal to the number of spectra). As an example we export below the
spectrum 1 and 3 to one file and spectra 2 and 4 to another.

```{r export-twofiles}
fls <- c(tempfile(), tempfile())
export(sps, MsBackendMzR(), file = fls[c(1, 2, 1, 2)])
```

A more realistic use case for mzML export would be to export MS data after
processing, such as smoothing (using the `smooth` function) and centroiding
(using the `pickPeaks` function) of raw profile-mode MS data.


## Changing backends

In the previous sections we learned already that a `Spectra` object can use
different backends for the actual data handling. It is also possible to
change the backend of a `Spectra` to a different one with the `setBackend`
function. We could for example change the (`MsBackendMzR`) backend of the
`sps_sciex` object to a `MsBackendDataFrame` backend to enable use of the data
even without the need to keep the original mzML files. Below we change the
backend of `sps_sciex` to the in-memory `MsBackendDataFrame` backend.

```{r setbackend}
print(object.size(sps_sciex), units = "Mb")
sps_sciex <- setBackend(sps_sciex, MsBackendDataFrame())
sps_sciex
```

With the call the full peak data was imported from the original mzML files into
the object. This has obviously an impact on the object's size, which is now much
larger than before.

```{r memory-after-import}
print(object.size(sps_sciex), units = "Mb")
```

The `dataStorage` spectrum variable has now changed, while `dataOrigin` still
keeps the information about the originating files:

```{r new-datastorage}
head(dataStorage(sps_sciex))
head(basename(dataOrigin(sps_sciex)))
```

## Parallel processing notes

Most functions on `Spectra` support (and use) parallel processing out of the
box. Peak data access and manipulation methods perform by default parallel
processing on a per-file basis (i.e. using the `dataStorage` variable as
splitting factor). `Spectra` uses `r Biocpkg("BiocParallel")` for
parallel processing and all functions use the default registered parallel
processing setup of that package.


# Backends

Backends allow to use different *backends* to store mass spectrometry data while
providing *via* the `Spectra` class a unified interface to use that data. The
`Spectra` package defines a set of example backends but any object extending the
base `MsBackend` class could be used instead. The default backends are:

- `MsBackendDataFrame`: the mass spectrometry data is stored (in-memory) in a
  `DataFrame`. Keeping the data in memory guarantees high performance but has
  also, depending on the number of mass peaks in each spectrum, a much higher
  memory footprint.

- `MsBackendMzR`: this backend keeps only general spectra variables in memory
  and relies on the `r Biocpkg("mzR")` package to read mass peaks (m/z and
  intensity values) from the original MS files on-demand.

- `MsBackendHdf5Peaks`: similar to `MsBackendMzR` this backend reads peak data
  only on-demand from disk while all other spectra variables are kept in
  memory. The peak data are stored in Hdf5 files which guarantees scalability.

All of the above mentioned backends support changing all of their their spectra
variables, **except** the `MsBackendMzR` that does not support changing m/z or
intensity values for the mass peaks.

With the example below we load the data from a single mzML file and use a
`MsBackendHdf5Peaks` backend for data storage. The `hdf5path` parameter allows
us to specify the storage location of the HDF5 file.

```{r hdf5}
library(msdata)
fl <- proteomics(full.names = TRUE)[5]

sps_tmt <- Spectra(fl, backend = MsBackendHdf5Peaks(), hdf5path = tempdir())
head(basename(dataStorage(sps_tmt)))
```

A (possibly incomplete) list of R packages providing additional backends that
add support for additional data types or storage options is provided below:

- `r BiocStyle::Biocpkg("MsBackendMgf")`: support for import/export of mass
  spectrometry files in mascot generic format (MGF).
- `r BiocStyle::Biocpkg("MsBackendMassbank")`: support for import/export and
  access of MassBank files/databases.
- [MsBackendHmdb](https://github.com/rformassspectrometry/MsBackendHmdb):
  support for import of MS2 spectra from files in the xml-file format of the
  Human Metabolome Database (HMDB).
- [MsBackendMsp](https://github.com/rformassspectrometry/MsBackendMsp): import
  MS2 spectra from files in MSP format.


# Session information

```{r si}
sessionInfo()
```

# References
